---
title: "EncPos"
author: "Fernanda Lang Schumacher"
date: "23/11/2020"
header-includes:
   - \usepackage{verbatim}
   - \usepackage{multicol}
output: 
  beamer_presentation:
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, fig.width = 6, fig.height = 4, fig.align = 'center')
library(tidyverse,quietly = T)
library(lme4,quietly = T)
library(nlme,quietly = T)
library(skewlmm,quietly = T)
library(gridExtra,quietly = T)
```

## Motivation: sleepstudy data

<!-- https://garrettgman.github.io/rmarkdown/beamer_presentation_format.html -->
- The average reaction time per day for subjects was evaluated by Gregory et al. (2003) in a sleep deprivation study. 
- On day 0 the subjects had their normal amount of sleep and starting that night they were restricted to 3 hours of sleep per night for 9 days, and the reaction time basead on a series of tests was measured on each day for each subject. 
- The data are avaliable at the R package \emph{lme4}.

## Motivation: sleepstudy data

```{r data, out.width = '85%'}
data(sleepstudy)

ggplot(sleepstudy,aes(x=Days,y=Reaction,group=Subject)) + geom_line(alpha=.3) + 
  geom_point(alpha=.3) +stat_summary(aes(group = 1),geom = "line", fun= mean, colour="blue") +
  theme_minimal()
```

## Introduction

- Bullet 1
- Bullet 2
- Bullet 3

## Linear mixed models

## Fitting a LMM to the sleepstudy dataset

<!-- ctrl + alt + i -->
```{r fit1, echo=TRUE}
fitlme <- lme(Reaction~Days,data=sleepstudy,
            random=~Days|Subject)
```

```{r fit1plot, out.width = '70%'}
grid.arrange(ggplot(data = nlme::ranef(fitlme),aes(x=`(Intercept)`)) + geom_density() + theme_minimal(),
ggplot(data = nlme::ranef(fitlme),aes(sample=`(Intercept)`)) +  geom_qq() +geom_qq_line() + theme_minimal(),
ggplot(data = nlme::ranef(fitlme),aes(x=`Days`)) + geom_density() + theme_minimal(),
ggplot(data = nlme::ranef(fitlme),aes(sample=`Days`)) +  geom_qq() +geom_qq_line() + theme_minimal())
```

## Model formulation

## Within-subject dependence structures

## Tools for model evaluation

## The \texttt{R} package \emph{skewlmm}

- The package \emph{skewlmm} implemets an EM-type algorithm in \texttt{R} using S3 class, containing methods for estimating and predicting the SMSN-LMM.

- It has an user-friendly interface with generic \texttt{R} functions `print`, `summary`, `plot`, `fitted`, `residuals` and `predict` implemented. 

- The main functions in the package are the `smsn.lmm()` and `smn.lmm()` functions, which estimates the parameter of a SMSN-LMM and a SMN-LMM, respectively.

## The \texttt{R} package \emph{skewlmm}

The basic syntax of these functions is as follows:
```{r sintax, eval=FALSE, echo=TRUE}
smsn.lmm(data, formFixed, groupVar, formRandom, 
         depStruct, distr, ...)
smn.lmm(data, formFixed, groupVar, formRandom, 
        depStruct, distr, ...)
```
where

- `data`: A data frame containing all the variables to be used in the model.

- `formFixed`: A two-sided linear formula object describing the fixed effects part of the model.

- `groupVar`: A character containing the name of the variable which represents the subjects or groups in data.

***
- `formRandom`: A one-sided linear formula object describing the random effects part of the model.

- `depStruct`: A character indicating which dependence structure should be used.

- `distr`: A character indicating which distribution should be used.\pause

Some other useful arguments:

- `timeVar`: A character containing the name of the variable which represents the time in data.

- `initialValues`: A named list containing initial parameter values.

- `pAR`: The order of the autoregressive process that should be used (if depStruct="ARp").

***

The functions return an object of the class `SMSN` and `SMN`, respectively, containing a list of elements, and the following methods/functions are available to these classes:

<!-- \begin{multicols}{2} -->
- `print`

- `summary`

- `fitted`

- `plot`

- `predict`

- `residuals`

- `ranef`

- `acfresid`

- `healy.plot`

- `lr.test`

- `mahalDist`
<!-- \end{multicols} -->

## Example: sleepstudy

\tiny
```{r fit0, echo=T}
fit0<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,
              formRandom = ~Days,groupVar = "Subject",quiet = T)
fit0
```
***
\tiny
```{r fitskew0, echo=T}
fitskew0<-smsn.lmm(data=sleepstudy,formFixed = Reaction~Days,
              formRandom = ~Days,groupVar = "Subject",quiet = T)
fitskew0
```

## Changing the distribution
\scriptsize
```{r fit2, echo=T}
fit1<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,distr = 't',
              formRandom = ~Days,groupVar = "Subject",quiet = T)
fitskew1<-smsn.lmm(data=sleepstudy,formFixed = Reaction~Days,distr = 'st',
                  formRandom = ~Days,groupVar = "Subject",quiet = T)
fit2<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,distr = 'sl',
              formRandom = ~Days,groupVar = "Subject",quiet = T)
fitskew2<-smsn.lmm(data=sleepstudy,formFixed = Reaction~Days,distr = 'ssl',
                  formRandom = ~Days,groupVar = "Subject",quiet = T)
fit3<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,distr = 'cn',
              formRandom = ~Days,groupVar = "Subject",quiet = T)
fitskew3<-smsn.lmm(data=sleepstudy,formFixed = Reaction~Days,distr = 'scn',
                  formRandom = ~Days,groupVar = "Subject",quiet = T)
```

## Comparing the fitted models
```{r fit3, echo=F}
data.frame(distr=c('norm','sn','t','st','sl','ssl','cn','scn'),
rbind(unlist(fit0$criteria), 
      unlist(fitskew0$criteria),
      unlist(fit1$criteria), 
      unlist(fitskew1$criteria),
      unlist(fit2$criteria), 
      unlist(fitskew2$criteria),
      unlist(fit3$criteria), 
      unlist(fitskew3$criteria))) %>% knitr::kable(digits = 1)
```
## Healy plot
\small
```{r healy1, echo=T,out.width='85%'}
grid.arrange(healy.plot(fit0),
            healy.plot(fit1),
            healy.plot(fit2),
            healy.plot(fit3))
```

## LR test for the SCN model
```{r fit4, echo=T}
lr.test(fitskew3,fit3)
```


## Computing the ACF
```{r fit5, echo=T}
acfresid(fit3)
```
## Plotting the ACF
```{r fit6, echo=T,out.width='85%'}
plot(acfresid(fit3,calcCI = T,maxLag = 6))
```

## Fitting an AR(p)-SMN-LMM
\small
```{r fitAR, echo=T}
#sl
fit2ar1<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,
                 formRandom = ~Days,groupVar = "Subject",
                 distr="sl",depStruct = "ARp",pAR=1,quiet=T)
fit2ar2<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,
                 formRandom = ~Days,groupVar = "Subject",
                 distr="sl",depStruct = "ARp",pAR=2,quiet=T)
#cn
fit3ar1<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,
                 formRandom = ~Days,groupVar = "Subject",
                 distr="cn",depStruct = "ARp",pAR=1,quiet=T)
fit3ar2<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,
                 formRandom = ~Days,groupVar = "Subject",
                 distr="cn",depStruct = "ARp",pAR=2,quiet=T)
```

## Fitting a DEC-SMN-LMM
\small
```{r fitDEC, echo=T}
#sl
fit2dec<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,
                 formRandom = ~Days,groupVar = "Subject",
                 distr="sl",depStruct = "DEC",quiet=T,
                 timeVar = 'Days')
#cn
fit3dec<-smn.lmm(data=sleepstudy,formFixed = Reaction~Days,
                 formRandom = ~Days,groupVar = "Subject",
                 distr="cn",depStruct = "DEC",quiet=T,
                 timeVar = 'Days')
```
Since the data are equally spaced and sorted by time, the use of `timeVar` in here is optional (the function will use the position if `timeVar` is not provided).

## Comparing the fitted models
```{r fitARDEC, echo=F}
data.frame(distr=rep(c('sl','cn'),each=4),depStruct=rep(c('CI','AR1','AR2','DEC'),2),
rbind(unlist(fit2$criteria), 
      unlist(fit2ar1$criteria), 
      unlist(fit2ar2$criteria),
      unlist(fit2dec$criteria), 
      unlist(fit3$criteria), 
      unlist(fit3ar1$criteria),
      unlist(fit3ar2$criteria), 
      unlist(fit3dec$criteria))) %>% knitr::kable(digits = 1)
```
## Healy plot
\small
```{r healy2, echo=T,out.width='85%'}
grid.arrange(healy.plot(fit2ar1),
            healy.plot(fit3ar1),
            healy.plot(fit3ar2),ncol=2)
```

## LR test
\small
```{r lrtest, echo=T}
lr.test(fit3ar1,fit3ar2)
```

## ACF plot for AR(1)-SL-LMM
```{r acf1, echo=T,out.width='85%'}
plot(acfresid(fit2ar1,calcCI = T,maxLag = 6))
```

## ACF plot for AR(1)-CN-LMM
```{r acf2, echo=T,out.width='85%'}
plot(acfresid(fit3ar1,calcCI = T,maxLag = 6))
```

## Fitted models
\scriptsize
```{r print1, echo=T}
#sl
cbind(fit2ar1$theta,fit2ar1$std.error) %>% 
  knitr::kable(digits = 2,col.names = c('estimate','std.error'))
```

***
\scriptsize
```{r print2, echo=T}
#cn
cbind(fit3ar1$theta,fit3ar1$std.error) %>% 
  knitr::kable(digits = 2,col.names = c('estimate','std.error'))
```


## Mahalanobis distance
\scriptsize
```{r mahal1, echo=T,out.width='80%', fig.height=3}
grid.arrange(plot(mahalDist(fit2ar1),fitobject = fit2ar1,nlabels = 0)+
               ggtitle('AR(1)-SL-LMM'),
             plot(mahalDist(fit3ar1),fitobject = fit3ar1,nlabels = 0)+
               ggtitle('AR(1)-CN-LMM'),
             ncol=2)
```

## Mahalanobis distance versus $\hat{u}$
\scriptsize
```{r mahal2, echo=T,out.width='80%', fig.height=3}
grid.arrange(qplot(mahalDist(fit2ar1),fit2ar1$uhat)+theme_minimal()+
               ylab("uhat")+xlab("Mahalanobis distance")+
               ggtitle('AR(1)-SL-LMM'),
             qplot(mahalDist(fit3ar1),fit3ar1$uhat)+theme_minimal()+
               ylab("uhat")+xlab("Mahalanobis distance")+
               ggtitle('AR(1)-CN-LMM'),
             ncol=2)
```


## Plot fit object
\scriptsize
```{r extra1, echo=T, out.width='85%', fig.height=3}
grid.arrange(plot(fit2ar1,type = "normalized")+ggtitle('AR(1)-SL-LMM'),
             plot(fit3ar1,type = "normalized")+ggtitle('AR(1)-CN-LMM'),
             ncol=2)
```


## Prediction
\scriptsize
```{r pred1, echo=T}
tail(sleepstudy,n=3)
predData <- data.frame(Reaction=NA,Days=10:11,Subject="372")
predict(fit2ar1,newData = predData)
predict(fit3ar1,newData = predData)
```

## References

- Gregory Belenky, Nancy J. Wesensten, David R. Thorne, Maria L. Thomas, Helen C. Sing, Daniel P. Redmond, Michael B. Russo and Thomas J. Balkin (2003) Patterns of performance degradation and restoration during sleep restriction and subsequent recovery: a sleep dose-response study. Journal of Sleep Research 12, 1–12.

